---
title: "OLE Analysis Workflow"
author: "Lei Shi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OLE Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(rdborrow)
```


## OLE phase

This vignette demonstrates the open-label extension (OLE) phase analysis
workflow using the difference-in-differences (DID) and synthetic control
method (SCM) estimators proposed in
[Zhou et al. (2024)](https://doi.org/10.1080/10543406.2024.2444222)
for estimating long-term treatment effects when the control group
switches to treatment.

### 1 DID methods

#### 1.1 IPW
```{r message=FALSE, warning=FALSE}
bootstrap_obj <- setup_bootstrap(
  replicates = 50,
  bootstrap_CI_type = "perc"
)

method_DID_obj <- setup_method_DID(
  method_name = "IPW",
  bootstrap_flag = TRUE,
  bootstrap_obj = bootstrap_obj,
  model_form_piS = "S ~ x1 + x2 + x3 + x4 + x5",
  model_form_piA = "A ~ x1 + x2 + x3 + x4 + x5"
)

analysis_OLE_obj <- setup_analysis_OLE(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2", "y3", "y4"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  T_cross = 2,
  method_OLE_obj = method_DID_obj
)

res <- run_analysis(analysis_OLE_obj)
```

#### 1.2 AIPW
```{r message=FALSE, warning=FALSE}
bootstrap_obj <- setup_bootstrap(
  replicates = 50,
  bootstrap_CI_type = "perc"
)

model_form_mu <- c(
  "y1 ~ x1 + x2 + x3 + x4 + x5",
  "y2 ~ x1 + x2 + x3 + x4 + x5",
  "y3 ~ x1 + x2 + x3 + x4 + x5",
  "y4 ~ x1 + x2 + x3 + x4 + x5"
)

method_DID_obj <- setup_method_DID(
  method_name = "AIPW",
  bootstrap_flag = TRUE,
  bootstrap_obj = bootstrap_obj,
  model_form_piS = "S ~ x1 + x2 + x3 + x4 + x5",
  model_form_piA = "A ~ x1 + x2 + x3 + x4 + x5",
  model_form_mu0_ext = model_form_mu
)

analysis_OLE_obj <- setup_analysis_OLE(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2", "y3", "y4"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  T_cross = 2,
  method_OLE_obj = method_DID_obj
)

res <- run_analysis(analysis_OLE_obj)
```


#### 1.3 OR
```{r message=FALSE, warning=FALSE}
bootstrap_obj <- setup_bootstrap(
  replicates = 50,
  bootstrap_CI_type = "perc"
)

model_form_mu <- c(
  "y1 ~ x1 + x2 + x3 + x4 + x5",
  "y2 ~ x1 + x2 + x3 + x4 + x5",
  "y3 ~ x1 + x2 + x3 + x4 + x5",
  "y4 ~ x1 + x2 + x3 + x4 + x5"
)

method_DID_obj <- setup_method_DID(
  method_name = "OR",
  bootstrap_flag = TRUE,
  bootstrap_obj = bootstrap_obj,
  model_form_mu0_ext = model_form_mu,
  model_form_mu0_rct = model_form_mu,
  model_form_mu1_rct = model_form_mu
)

analysis_OLE_obj <- setup_analysis_OLE(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2", "y3", "y4"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  T_cross = 2,
  method_OLE_obj = method_DID_obj
)

res <- run_analysis(analysis_OLE_obj)
res
```


### 2 SCM method: with parallel computing
```{r}
bootstrap_obj <- setup_bootstrap(
  replicates = 50,
  bootstrap_CI_type = "perc"
)

method_SCM_obj <- setup_method_SCM(
  method_name = "SCM",
  bootstrap_flag = TRUE,
  bootstrap_obj = bootstrap_obj,
  lambda.min = 0,
  lambda.max = 1e-3,
  nlambda = 10,
  parallel = "no",
  ncpus = 1
)

analysis_OLE_obj <- setup_analysis_OLE(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2", "y3", "y4"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  T_cross = 2,
  method_OLE_obj = method_SCM_obj
)

run_analysis(analysis_OLE_obj)
```


### 3 SCM method: without parallel computing
```{r}
bootstrap_obj <- setup_bootstrap(
  replicates = 50,
  bootstrap_CI_type = "perc"
)

method_SCM_obj <- setup_method_SCM(
  method_name = "SCM",
  bootstrap_flag = TRUE,
  bootstrap_obj = bootstrap_obj,
  lambda.min = 0,
  lambda.max = 1e-3,
  nlambda = 10,
  parallel = "no"
)

analysis_OLE_obj <- setup_analysis_OLE(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2", "y3", "y4"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  T_cross = 2,
  method_OLE_obj = method_SCM_obj
)

run_analysis(analysis_OLE_obj)
```

## References

- Zhou X, Pang H, Drake C, Burger HU, Zhu J (2024). "Estimating treatment effect in randomized trial after control to treatment crossover using external controls." *Journal of Biopharmaceutical Statistics*. doi: [10.1080/10543406.2024.2444222](https://doi.org/10.1080/10543406.2024.2444222).
- Shi L, Pang H, Chen C, Zhu J (2025). "rdborrow: an R package for causal inference incorporating external controls in randomized controlled trials with longitudinal outcomes." *Journal of Biopharmaceutical Statistics*, 35(6), 1043-1066. doi: [10.1080/10543406.2025.2489283](https://doi.org/10.1080/10543406.2025.2489283).
