---
title: "Adding a new method"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdborrow)
```

This article is for developers who want to add a borrowing estimator to
rdborrow. Every method is an S4 class with its own `estimate()` method.
`run_analysis()` and `run_simulation()` reach the estimator through S4
dispatch, so a new method never changes the logic of existing code. Adding
one to the package does touch a few existing files to register it (the
method lists in the docs, `NAMESPACE`, `NEWS.md`, `_pkgdown.yml`); the
checklist below lists every one.

## How a method plugs in

Every method object inherits from one of two parent classes, and that
parent decides which analysis it can be used in:

```
method_obj
├── method_primary_obj  -> setup_analysis_primary()
│   └── method_weighting_obj    ec_ipw(), ec_aipw()
└── method_OLE_obj      -> setup_analysis_OLE()
    ├── method_DID_obj          did_ec_ipw(), did_ec_aipw(), did_ec_or()
    └── method_SCM_obj          scm()
```

`method_obj` carries the slots every method shares: `method_name`,
`bootstrap`, and `bootstrap_ci_type`. Your class adds slots for its own
settings, such as model formulas or a borrowing weight.

`run_analysis()` then does one thing. It collects the column names stored in
the analysis object and calls:

```r
estimate(
  method, data, outcomes, treatment, trial_status, covariates,
  alpha, quiet
)
```

OLE analyses also pass `T_cross`.

## The `estimate()` contract

### What `estimate()` receives

`data` is the user's data frame, with the user's column names, holding
trial participants and external controls. `outcomes`, `treatment`,
`trial_status`, and `covariates` are the column names given to
`setup_analysis_*()`.

The built-in methods immediately standardize the data with the internal
helper `.build_analysis_df()`. It returns the outcome columns, then the trial
status column renamed to `S`, the treatment column renamed to `A`, then the
covariates. Here the trial status and treatment columns start out as
`in_trial` and `arm`:

```{r}
trial_data <- SyntheticData
names(trial_data)[names(trial_data) == "S"] <- "in_trial"
names(trial_data)[names(trial_data) == "A"] <- "arm"

df <- rdborrow:::.build_analysis_df(
  trial_data,
  outcomes = c("y1", "y2"), treatment = "arm", trial_status = "in_trial",
  covariates = c("x1", "x2")
)
head(df, 3)
```

Do the same. The shared bootstrap helper stratifies on columns literally
named `S` and `A`, and user formulas are rewritten to match: for example
`ec_ipw()` replaces the left-hand side of `ps_formula` with `S`.

Validate anything that depends on the analysis inside `estimate()`, not in
the constructor, because the constructor can't see the data. For example,
`ec_aipw()` checks there that it has one outcome formula per outcome.

### What `estimate()` returns

The analysis and simulation code expects a specific shape.

| Analysis | Return value |
|---|---|
| Primary, sandwich SE | `list(results, borrow_weight)`, where `results` is a data frame with one row per outcome (`tau1`, `tau2`, ...) and columns `point_estimates`, `standard_deviation`, `lower_CI_normal`, `upper_CI_normal` |
| Primary, bootstrap | The same, but `standard_deviation` is the bootstrap standard deviation and the CI columns are `lower_CI_boot`, `upper_CI_boot` instead of the `*_normal` pair |
| OLE | A data frame with one row per post-crossover outcome and columns `point_estimates`, `lower_CI_boot`, `upper_CI_boot` |

A primary method doesn't need a sandwich variance. If it only supports the
bootstrap, return just the bootstrap shape.

`run_simulation()` keeps the last row (the final time point) from each
replicate. To compute coverage, type I error, and power, it uses the
`*_boot` confidence interval columns if they exist and the `*_normal`
columns otherwise. Keep these column names exactly as shown. For the same
reason, `setup_simulation_*()` takes `true_effect` as a single number: the
effect at the final visit.

## A toy method, end to end

The fastest way to see the contract in action is to define a method outside
the package. The dummy estimator below, `hamburger()`, is deliberately named
so that nobody mistakes it for a real method. It pools trial controls and
external controls with a fixed weight `w` and makes no covariate adjustment:

$$
\hat\tau = \bar Y_{11} - \left[(1 - w)\,\bar Y_{10} + w\,\bar Y_{00}\right]
$$

Here $\bar Y_{11}$, $\bar Y_{10}$, and $\bar Y_{00}$ are the outcome means for
trial treated, trial control, and external control patients.

First, the class and a validated constructor. `bootstrap` and
`bootstrap_ci_type` are slots inherited from `method_obj`:

```{r}
.hamburger_method <- setClass(
  "hamburger_method",
  contains = "method_primary_obj",
  slots = c(weight = "numeric"),
  prototype = list(method_name = "Hamburger", weight = 0.5)
)

hamburger <- function(weight = 0.5, bootstrap = NULL,
                      bootstrap_ci_type = NULL) {
  checkmate::assert_number(weight, lower = 0, upper = 1)
  checkmate::assert_int(bootstrap, lower = 2, null.ok = TRUE)
  if (!is.null(bootstrap) && is.null(bootstrap_ci_type)) {
    bootstrap_ci_type <- "perc"
  }
  .hamburger_method(
    method_name = "Hamburger", weight = weight,
    bootstrap = bootstrap, bootstrap_ci_type = bootstrap_ci_type
  )
}
```

Next, the estimator. Following the package convention, one internal
function, `.hamburger_core()`, computes the point estimate. The bootstrap
statistic resamples rows and calls the same function, so the point estimate
is defined in exactly one place:

```{r}
.hamburger_core <- function(Y, S, A, weight) {
  groups <- list(
    trt = S == 1 & A == 1,
    ctrl = S == 1 & A == 0,
    ext = S == 0
  )
  means <- lapply(groups, \(g) colMeans(Y[g, , drop = FALSE]))
  vars <- lapply(groups, \(g) apply(Y[g, , drop = FALSE], 2, var) / sum(g))

  list(
    tau = means$trt - (1 - weight) * means$ctrl - weight * means$ext,
    sd_tau = sqrt(vars$trt + (1 - weight)^2 * vars$ctrl +
      weight^2 * vars$ext)
  )
}

.hamburger_boot_statistic <- function(data, indices, outcomes, weight) {
  d <- data[indices, , drop = FALSE]
  Y <- as.matrix(d[, outcomes, drop = FALSE])
  .hamburger_core(Y, d$S, d$A, weight)$tau
}
```

`estimate()` standardizes the data, computes the point estimate, and then
attaches either a closed-form standard error or bootstrap intervals. Inside
the package you would call `.build_analysis_df()` and `.run_bootstrap()`
directly; outside it they need `rdborrow:::`.

```{r}
setMethod("estimate", "hamburger_method", function(method, data, outcomes,
                                                   treatment, trial_status,
                                                   covariates, alpha = 0.05,
                                                   quiet = TRUE) {
  df <- rdborrow:::.build_analysis_df(
    data, outcomes, treatment, trial_status, covariates
  )
  Y <- as.matrix(df[, outcomes, drop = FALSE])
  core <- .hamburger_core(Y, df$S, df$A, method@weight)
  labels <- paste0("tau", seq_along(outcomes))

  if (is.null(method@bootstrap)) {
    cutoff <- qnorm(1 - alpha / 2)
    results <- data.frame(
      point_estimates = core$tau,
      standard_deviation = core$sd_tau,
      lower_CI_normal = core$tau - cutoff * core$sd_tau,
      upper_CI_normal = core$tau + cutoff * core$sd_tau,
      row.names = labels
    )
  } else {
    boot_res <- rdborrow:::.run_bootstrap(
      df = df, statistic = .hamburger_boot_statistic,
      n_estimates = length(outcomes), bootstrap = method@bootstrap,
      bootstrap_ci_type = method@bootstrap_ci_type, alpha = alpha,
      outcomes = outcomes, weight = method@weight
    )
    results <- data.frame(
      point_estimates = core$tau,
      standard_deviation = boot_res$sd_boot,
      lower_CI_boot = boot_res$lower_ci,
      upper_CI_boot = boot_res$upper_ci,
      row.names = labels
    )
  }

  list(results = results, borrow_weight = method@weight)
})
```

`.run_bootstrap()` takes the standardized data frame, the statistic, the
number of estimates it returns, the number of replicates, the CI type, and
`alpha`. Any further named arguments (here `outcomes` and `weight`) are
passed through to the statistic. It resamples within each trial status and
treatment group and returns a list with `lower_ci`, `upper_ci`, and
`sd_boot`, one value per estimate.

That is the whole method. It runs through the standard workflow without
any changes to rdborrow:

```{r}
analysis <- setup_analysis_primary(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  method_weighting_obj = hamburger(weight = 0.5)
)
run_analysis(analysis)
```

With `bootstrap` set, the same call returns bootstrap intervals:

```{r}
analysis@method_obj <- hamburger(weight = 0.5, bootstrap = 200)
set.seed(1)
run_analysis(analysis)
```

Swapping in `ec_ipw()` with the same fixed weight shows how much the
covariate adjustment changes the answer:

```{r}
analysis@method_obj <- ec_ipw(
  ps_formula = "S ~ x1 + x2 + x3 + x4 + x5",
  weight = 0.5
)
run_analysis(analysis)
```

For the same reason, `hamburger()` can go straight into the
`method_obj_list` of `setup_simulation_primary()` and be compared against
the built-in methods (see `vignette("primary_simulation_workflow")`).

## Moving the method into the package

Defining a method outside the package works for prototyping. To add it to
rdborrow itself, use `R/ec_ipw.R` as the template.

1. **Create `R/{name}.R`.** Start it with `#' @include method_class.R`.
   rdborrow uses a `Collate` field, so `devtools::document()` needs this
   tag to load the parent classes first. Functions from the packages and
   `stats` functions that rdborrow already imports (see `R/package.R`) can
   be used directly. For anything else, add an `@importFrom` there or call
   it as `pkg::fun()`.
2. **Define the class and constructor.** Validate every argument with
   checkmate, including `checkmate::assert_int(bootstrap, lower = 2,
   null.ok = TRUE)` (drop `null.ok` if the method is bootstrap only).
   Resolve `bootstrap_ci_type` to `"perc"` when `bootstrap` is set.
   Document the constructor with roxygen, `@export` it, and give it
   runnable `@examples`.
3. **Write the `estimate()` method.** Tag it with `#' @rdname estimate`.
   Start with `.build_analysis_df()`, and put any validation that depends
   on the analysis, such as one formula per outcome, here.
4. **Split the internals into three functions:**
    - `.{name}_core()`: the single source of the point estimate.
    - `.{name}_se()`: the sandwich variance, if the method has a closed
      form. Skip it for a bootstrap-only method, as the DID methods and
      `scm()` do.
    - `.{name}_boot_statistic(data, indices, ...)`: a thin wrapper that
      resamples rows and calls `.{name}_core()`. Pass it to the shared
      `.run_bootstrap()`, as in the toy method above. Don't write a new
      bootstrap loop.

    Describe internal functions with `#'` comments that end in `@noRd`, as
    `R/ec_ipw.R` does, so they don't get a help page.
5. **Lock the numbers with a test.** Add
   `tests/testthat/test-full_pipeline_{name}.R` that runs the method on
   `SyntheticData` through `setup_analysis_*()` and `run_analysis()`. Check
   the point estimates against an independent reference, such as the paper's
   code or a hand calculation, rather than against the new function's own
   output. Call `set.seed()` before any bootstrap. Model refits on small
   resamples can warn (for example, rank-deficient fits on `SyntheticData`);
   expect those warnings explicitly with `expect_warning()`.
6. **Register the method.** Add it to:
    - the method lists in `?run_analysis` (`R/run_analysis.R`),
      `?run_simulation` (`R/run_simulation.R`), and
      `?setup_analysis_primary` (`R/analysis_primary_class.R`) or
      `?setup_analysis_OLE` (`R/analysis_OLE_class.R`)
    - the "Borrowing methods" section of `_pkgdown.yml`
    - `NEWS.md`

    `devtools::document()` then updates `NAMESPACE` and the matching
    `man/*.Rd` files. These registration edits are the only changes to
    existing files.
7. **Validate:**

    ```r
    devtools::document()
    styler::style_pkg()
    devtools::test()
    pkgdown::check_pkgdown()
    devtools::check()
    ```
