---
title: "Adding a new method"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdborrow)
```

This article is for developers who want to add a borrowing estimator to
rdborrow. Every method is an S4 class with its own `estimate()` method.
`run_analysis()` and `run_simulation()` reach the estimator through S4
dispatch, so a new method never requires editing existing code.

## How a method plugs in

Every method object inherits from one of two parent classes, and that
parent decides which analysis it can be used in:

```
method_obj
├── method_primary_obj  -> setup_analysis_primary()
│   └── method_weighting_obj    ec_ipw(), ec_aipw()
└── method_OLE_obj      -> setup_analysis_OLE()
    ├── method_DID_obj          did_ec_ipw(), did_ec_aipw(), did_ec_or()
    └── method_SCM_obj          scm()
```

`method_obj` carries the slots every method shares: `method_name`,
`bootstrap`, and `bootstrap_ci_type`. Your class adds slots for its own
settings, such as model formulas or a borrowing weight.

`run_analysis()` then does one thing. It collects the column names stored in
the analysis object and calls:

```r
estimate(
  method, data, outcomes, treatment, trial_status, covariates,
  alpha, quiet
)
```

OLE analyses also pass `T_cross`.

## The `estimate()` contract

Your `estimate()` method receives the full data frame (trial participants
and external controls) and the column names. The analysis and simulation
code expects it to return a specific shape.

| Analysis | Return value |
|---|---|
| Primary | `list(results, borrow_weight)`, where `results` is a data frame with one row per outcome (`tau1`, `tau2`, ...) and columns `point_estimates`, `standard_deviation`, and either `lower_CI_normal`/`upper_CI_normal` or, when bootstrapping, `lower_CI_boot`/`upper_CI_boot` |
| OLE | A data frame with one row per post-crossover outcome and columns `point_estimates`, `lower_CI_boot`, `upper_CI_boot` |

`run_simulation()` keeps the last row (the final time point) from each
replicate. To compute coverage, type I error, and power, it uses the
`*_boot` confidence interval columns if they exist and the `*_normal`
columns otherwise. Keep these column names exactly as shown.

## A toy method, end to end

The fastest way to see the contract in action is to define a method outside
the package. The dummy estimator below, `hamburger()`, is deliberately named
so that nobody mistakes it for a real method. It pools trial controls and
external controls with a fixed weight `w` and makes no covariate adjustment:

$$
\hat\tau = \bar Y_{11} - \left[(1 - w)\,\bar Y_{10} + w\,\bar Y_{00}\right]
$$

Here $\bar Y_{11}$, $\bar Y_{10}$, and $\bar Y_{00}$ are the outcome means for
trial treated, trial control, and external control patients.

First, the class and a validated constructor:

```{r}
.hamburger_method <- setClass(
  "hamburger_method",
  contains = "method_primary_obj",
  slots = c(weight = "numeric"),
  prototype = list(method_name = "Hamburger", weight = 0.5)
)

hamburger <- function(weight = 0.5) {
  checkmate::assert_number(weight, lower = 0, upper = 1)
  .hamburger_method(method_name = "Hamburger", weight = weight)
}
```

Next, the estimator. Following the package convention, one internal
function, `.hamburger_core()`, computes the point estimate, and
`estimate()` wraps it with a standard error and confidence interval:

```{r}
.hamburger_core <- function(Y, S, A, weight) {
  groups <- list(
    trt = S == 1 & A == 1,
    ctrl = S == 1 & A == 0,
    ext = S == 0
  )
  means <- lapply(groups, \(g) colMeans(Y[g, , drop = FALSE]))
  vars <- lapply(groups, \(g) apply(Y[g, , drop = FALSE], 2, var) / sum(g))

  list(
    tau = means$trt - (1 - weight) * means$ctrl - weight * means$ext,
    sd_tau = sqrt(vars$trt + (1 - weight)^2 * vars$ctrl +
      weight^2 * vars$ext)
  )
}

setMethod("estimate", "hamburger_method", function(method, data, outcomes,
                                                   treatment, trial_status,
                                                   covariates, alpha = 0.05,
                                                   quiet = TRUE) {
  Y <- as.matrix(data[, outcomes, drop = FALSE])
  core <- .hamburger_core(
    Y, data[[trial_status]], data[[treatment]], method@weight
  )
  cutoff <- qnorm(1 - alpha / 2)

  results <- data.frame(
    point_estimates = core$tau,
    standard_deviation = core$sd_tau,
    lower_CI_normal = core$tau - cutoff * core$sd_tau,
    upper_CI_normal = core$tau + cutoff * core$sd_tau,
    row.names = paste0("tau", seq_along(outcomes))
  )
  list(results = results, borrow_weight = method@weight)
})
```

That is the whole method. It runs through the standard workflow without
any changes to rdborrow:

```{r}
analysis <- setup_analysis_primary(
  data = SyntheticData,
  trial_status_col_name = "S",
  treatment_col_name = "A",
  outcome_col_name = c("y1", "y2"),
  covariates_col_name = c("x1", "x2", "x3", "x4", "x5"),
  method_weighting_obj = hamburger(weight = 0.5)
)
run_analysis(analysis)
```

Swapping in `ec_ipw()` with the same fixed weight shows how much the
covariate adjustment changes the answer:

```{r}
analysis@method_obj <- ec_ipw(
  ps_formula = "S ~ x1 + x2 + x3 + x4 + x5",
  weight = 0.5
)
run_analysis(analysis)
```

For the same reason, `hamburger()` can go straight into the
`method_obj_list` of `setup_simulation_primary()` and be compared against
the built-in methods (see `vignette("primary_simulation_workflow")`).

## Moving the method into the package

Defining a method outside the package works for prototyping. To add it to
rdborrow itself, use `R/ec_ipw.R` as the template.

1. **Create `R/{name}.R`.** Start it with `#' @include method_class.R`.
   rdborrow uses a `Collate` field, so `devtools::document()` needs this
   tag to load the parent classes first.
2. **Define the class and constructor.** Validate every argument with
   checkmate. Resolve `bootstrap_ci_type` to `"perc"` when `bootstrap` is
   set. Document the constructor with roxygen, `@export` it, and give it
   runnable `@examples`.
3. **Write the `estimate()` method.** Tag it with `#' @rdname estimate`.
4. **Split the internals into three functions:**
    - `.{name}_core()`: the single source of the point estimate.
    - `.{name}_se()`: the sandwich variance, if the method has a closed
      form. The DID and SCM methods don't and support bootstrap only.
    - `.{name}_boot_statistic(data, indices, ...)`: a thin wrapper that
      resamples rows and calls `.{name}_core()`. Pass it to the shared
      `.run_bootstrap()`, which stratifies by trial status and treatment.
      Don't write a new bootstrap loop.
5. **Lock the numbers with a test.** Add
   `tests/testthat/test-full_pipeline_{name}.R` that runs the method on
   `SyntheticData` through `setup_analysis_*()` and `run_analysis()`. Check
   the point estimates against an independent reference, such as the paper's
   code or a hand calculation, rather than against the new function's own
   output.
6. **Register the method.** Add it to the "Borrowing methods" section of
   `_pkgdown.yml`, to the method list in `?run_analysis`, and to `NEWS.md`.
7. **Validate:**

    ```r
    devtools::document()
    styler::style_pkg()
    devtools::test()
    pkgdown::check_pkgdown()
    devtools::check()
    ```

## Using an AI coding agent

The steps above are mechanical once the math is settled, which makes them a
good fit for an AI coding agent. [Developing with AI coding
agents](ai-coding-agents.html) shows how to hand this recipe to an agent and
how to check its work.
